---
title: 'Portfolio 4: The Alien Game'
output: html_document
date: "2025-04-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load in packages
```{r cars}
pacman::p_load(tidyverse,
        here,
        posterior,
        betafunctions,
        cmdstanr, # Stan need to be at least version 0.8.xxx
        brms,
        updog,
        tidybayes,
        extraDistr,
        cowplot,
        loo,
        ggplot2,
        patchwork,
        gridExtra,
        grid) 


```


# Setup
```{r}
# Set working directory
setwd(".")

# Load emperical data
df <- read.csv("AlienData.txt", header = TRUE)
df <- df %>% filter(condition == 1 & test == 0)

```


# Defining dangerous categorization function
```{r}
dangerous_rule <- function(session, F1_eyes, F2_legs, F3_body, F4_arms, F5_color) {
  if (session == 1) {
    if (F1_eyes + F3_body == 2) {
      dangerous = 1 
    } else {
      dangerous = 0
    }
  }
    
  else if (session == 2) {
    if (F4_arms == 1) {
      dangerous = 1
    } else {
      dangerous = 0
    }
  }
    
  else if (session == 3) {
    if (F4_arms + F5_color == 2) {
      dangerous = 1 
    } else {
      dangerous = 0
    }
  }
  return(dangerous)
}

```


# Defning a distance function (courtsey of Riccardo)
```{r}
distance <- function(vect1, vect2, w){
  return(sum(w * abs(vect1 - vect2)))
}
```


# Defining a similarity function (courtsey of Riccardo)
```{r}
similarity <- function(distance, c){
  return(exp(-c * distance))
}
```


# Defining a gcm (courtsey of Riccardo)
```{r}
### generative model ###
gcm <- function(w, c, obs, cat_one, quiet = TRUE) {
  # create an empty list to save probability of saying "1" for each trial
  r <- c()
  
  ntrials <- nrow(obs)
  
  for (i in 1:ntrials) {
    # If quiet is FALSE, print every ten trials
    if (!quiet && i %% 10 == 0) {
      print(paste("i =", i))
    }
    # if this is the first trial, or there any category with no exemplars seen yet, set the choice to random
    if (i == 1 || sum(cat_one[1:(i - 1)]) == 0 || sum(cat_one[1:(i - 1)]) == (i - 1)) {
      r <- c(r, .5)
    } else {
      similarities <- c()
      # for each previously seen stimulus assess distance and similarity
      for (e in 1:(i - 1)) {
        sim <- similarity(distance(obs[i, ], obs[e, ], w), c)
        similarities <- c(similarities, sim)
      }
      # Calculate prob of saying "1" by dividing similarity to 1 by the sum of similarity to 1 and to 2
      numerator <- mean(similarities[cat_one[1:(i - 1)] == 1])
      denominator <- mean(similarities[cat_one[1:(i - 1)] == 1]) + mean(similarities[cat_one[1:(i - 1)] == 0])
      r <- c(r, numerator / denominator)
    }
  }

  return(rbinom(ntrials, 1, r))
}
```


# Defining response function (courtsey of Riccardo with small adjustments)
```{r}
simulate_responses <- function(data, agent, w, c){
  
  # select our features
  features <- data %>%
    dplyr::select(c("F1_eyes", "F2_legs", "F3_body", "F4_arms", "F5_color"))
  
  dangerous <- data$dangerous
  
  # define w/weight
  if (w == "equal") {
    weight <- c(1,1,1,1,1)
  }
  
  # simulate responses
  response <- gcm(
    weight,
    c,
    features,
    dangerous
  )
  
  # Generate a temporary dataframe
  temp_simulated_responses <- data %>%
    mutate(
      response = response,
      correct = ifelse(dangerous == response, 1, 0),
      performance = cumsum(correct) / seq_along(correct),
      c = c,
      w = w,
      agent = agent)

  return(temp_simulated_responses)
}
```


# Simulation
```{r}
# Create simulated data frame
sim_data <- data.frame(stimulus = df$stimulus)

# Experiment details 
n <- 25 # in the empirical data, there is no subject 13 which means that the sample size is 25 instead of 26
session <- 3 
trials_per_session <- 96
trials_per_participant <- trials_per_session*session # 288 in total

sim_data$subject <- rep(1:n, each = trials_per_participant) # repeat each subject (1 to 25) for all 288 trials
sim_data$session <- rep(rep(1:session, each = trials_per_session), times = n) # each subject gets a session number 1 to 3 
sim_data$trial <- rep(1:trials_per_session, times = n*session) # repeat trial numbers 1 to 96 for each session

# Initialize empty vectors 
sim_data$dangerous <- NA
#sim_data$response <- NA # the simulate response function does not work if this column already exists
#sim_data$correct <- NA # the simulate response function does not work if this column already exists

# Simulate for multiple participants
for (i in 1:nrow(sim_data)) {
  # save the string of the 'stimulus' column for that row
  stimulus_string <- as.character(sim_data$stimulus[i])
    
  # get the individual characters and fill the corresponding feature columns
  sim_data$F1_eyes[i] <- as.numeric(substr(stimulus_string, 1, 1))
  sim_data$F2_legs[i] <- as.numeric(substr(stimulus_string, 2, 2))
  sim_data$F3_body[i] <- as.numeric(substr(stimulus_string, 3, 3))
  sim_data$F4_arms[i] <- as.numeric(substr(stimulus_string, 4, 4))
  sim_data$F5_color[i] <- as.numeric(substr(stimulus_string, 5, 5))
  
  # Categorizing dangerous aliens 
  sim_data$dangerous[i] <- dangerous_rule(sim_data$session[i], sim_data$F1_eyes[i], sim_data$F2_legs[i], sim_data$F3_body[i], sim_data$F4_arms[i], sim_data$F5_color[i])
}

# Initialize an empty dataframe to store results
all_simulated_responses <- data.frame()

# Loop through each subject
for (subj in unique(sim_data$subject)) {
  
  # Filter the data for the current subject
  sim_data_subj <- sim_data %>% 
    filter(subject == subj)
  
  # Simulate responses for this subject
  simulated_responses <- simulate_responses(data = sim_data_subj, agent = "equal weights", w = "equal", c = 1)
  
  # Combine the results
  all_simulated_responses <- bind_rows(all_simulated_responses, simulated_responses)
}

# Save as object
saveRDS(all_simulated_responses, "all_simulated_responses.rds")
saveRDS(simulated_responses, "last_sub_responses.rds")

```


# Load rdata
```{r}
# Load object
all_simulated_responses <- readRDS("all_simulated_responses.rds")
simulated_responses <- readRDS("last_sub_responses.rds")
```


# Only work with session 1
```{r}
#drop session 2 and 3
all_simulated_responses <- all_simulated_responses %>% 
  filter(session %in% c(1))
```


# Investigate the simulated response 
```{r}
# Mean of correct by participant
all_simulated_responses %>% 
  group_by(subject) %>% 
  summarize(mean(correct, na.rm = TRUE))

# Overall mean of correct
all_simulated_responses %>% 
  summarize(mean(correct, na.rm = TRUE))

# Performance over time 

## Add performance to the df
all_simulated_responses <- all_simulated_responses %>%
  group_by(subject) %>% 
  mutate(performance = cumsum(correct) / seq_along(correct))

## Per participant
ggplot(all_simulated_responses, aes(x = trial, y = performance, color = as.factor(subject))) +
  geom_line() +                
  geom_point() +               
  labs(title = "Performance per subject",
       x = "Trial",
       y = "Performance",
       color = "Subject") +    
  theme_minimal()  
```


# Limit to 5 participants 
```{r}
five_simulated_responses <- all_simulated_responses %>% 
  filter(subject %in% c(1, 2, 3,4,5))
```


# Loop over 5 participants and apply STAN code
```{r}
# Specify path to stan model and load 
file <- file.path("gcm_stan.stan")

model <- cmdstan_model(file, 
                     cpp_options = list(stan_threads = TRUE),
                     stanc_options = list("O1"))

# Get the unique subject IDs 
subject_ids <- unique(five_simulated_responses$subject)

# Store model fits 
model_fits <- list()

for (subj in subject_ids) {
  cat("Fitting model for subject:", subj, "\n")
  
  subj_data <- five_simulated_responses %>% filter(subject == subj)
  
  data <- list(
    ntrials = nrow(subj_data),
    nfeatures = 5,
    cat_one = subj_data$dangerous,
    y = subj_data$response,
    obs = as.matrix(subj_data[, c("F1_eyes", "F2_legs", "F3_body", "F4_arms", "F5_color")]),
    b = 0.5,
    w_prior_values = rep(1, 5),
    c_prior_values = c(0, 1)
  )
  
  fit <- model$sample(
    data = data,
    seed = 123,
    chains = 2, 
    parallel_chains = 2, 
    threads_per_chain = 2, 
    iter_warmup = 1000,
    iter_sampling = 1000,
    refresh = 0,
    max_treedepth = 20,
    adapt_delta = 0.99
  )
  
  model_fits[[as.character(subj)]] <- fit
}


# Print diagnostics 
for (subj in names(model_fits)) {
   cat("\n--- Diagnostics for Subject", subj, "---\n")
   print(model_fits[[subj]]$cmdstan_diagnose())
}

# Print and save diagnostics per subject (to be used later)
for (subj in names(model_fits)) {
  
  # Extract summary and convert to tibble
  diagnostics <- model_fits[[subj]]$summary() %>% as_tibble()
  
  # Assign subject name 
  assign(paste0("subject_", subj, "_diagnostics"), diagnostics)
  
  # Save the tibble as an RDS file
  saveRDS(diagnostics, paste0("subject_", subj, "_diagnostics.rds"))
  
  # Print the summary
  print(diagnostics)
}



# Plot prior-posterior updates and traceplots
for (subj in names(model_fits)) {
  
  # Extract posterior draws
  draws <- as_draws_df(model_fits[[subj]]$draws())

  
  # Prior-posterior update checks for w's
  w_prior_pos <- lapply(1:5, function(i) {
    ggplot(draws)+
      geom_density(aes(x = .data[[paste0("w[", i, "]")]]), alpha = 3, fill = "lightblue") +
      geom_density(aes(x = .data[[paste0("w_prior[", i, "]")]]), alpha = 3, fill = "pink") +
      geom_vline(xintercept = 0.2, size = 1) + 
      theme_classic() + 
      ggtitle(paste("Prior-Posterior Update for w[", i, "]", sep = "")) +
      xlab(paste0("w[", i, "]")) +
      ylab("Posterior Density")
  })
  
  # Prior-posterior update checks for c
  c_prior_pos <- ggplot(draws) +
    geom_density(aes(c), alpha = 3, fill = "lightblue") +
    geom_density(aes(c_prior), alpha = 3, fill = "pink") +
    geom_vline(xintercept = 1, size = 1) + 
    theme_classic() + 
    ggtitle("Prior-Posterior Update for c") +
    xlab("c") +
    ylab("Posterior Density")
  
  # Combine the prior-posterior update checks
  grid_plots <- c(w_prior_pos, list(c_prior_pos))
  grid.arrange(grobs = grid_plots, 
               ncol = 2, 
               top = paste("Prior-posterior update checks - Subject", subj),
               bottom = "Prior = pink, Posterior = blue")
  
  # Traceplots for w's
  w_plots <- lapply(1:5, function(i) {
    ggplot(draws, aes(x = .iteration, y = .data[[paste0("w[", i, "]")]], 
                      group = .chain, color = .chain)) +
      geom_line(alpha = 0.5) +
      theme_classic() +
      ggtitle(paste("w[", i, "]", sep = ""))
  })
  
  # Traceplot for c 
  c_plot <- ggplot(draws, aes(x = .iteration, y = c, group = .chain, color = .chain)) +
    geom_line(alpha = 0.5) +
    theme_classic() +
    ggtitle("c")
  
  # Combine the traceplots
  grid_plots <- c(w_plots, list(c_plot))
  grid.arrange(grobs = grid_plots, ncol = 2, top = paste("Traceplots - Subject", subj))
}


```


# Combine the diagnostic putputs
```{r}
# Get the data frames into the right shape
subj_1 <- subject_1_diagnostics %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)

subj_2 <- subject_2_diagnostics %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)

subj_3 <- subject_3_diagnostics %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)

subj_4 <- subject_4_diagnostics %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)

subj_5 <- subject_5_diagnostics %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)


# Combine the data frames
combined_data <- bind_rows(subj_1, subj_2, subj_3, subj_4, subj_5)


# Calculate the mean for each variable across participants
mean_data <- combined_data %>%
  group_by(variable) %>%
  summarise(mean_value = mean(mean),
            median_value = median(mean),
            sd_value = sd(mean),
            mad_value = mad(mean))
mean_data
```


# The emperical data
```{r}
# Create columns of features

# Simulate for multiple participants
for (i in 1:nrow(df)) {
  # save the string of the 'stimulus' column for that row
  stimulus_string <- as.character(df$stimulus[i])
    
  # get the individual characters and fill the corresponding feature columns
  df$F1_eyes[i] <- as.numeric(substr(stimulus_string, 1, 1))
  df$F2_legs[i] <- as.numeric(substr(stimulus_string, 2, 2))
  df$F3_body[i] <- as.numeric(substr(stimulus_string, 3, 3))
  df$F4_arms[i] <- as.numeric(substr(stimulus_string, 4, 4))
  df$F5_color[i] <- as.numeric(substr(stimulus_string, 5, 5))
  
}
```


# Limit to session 1
```{r}
#drop session 2 and 3
df <- df %>% 
  filter(session %in% c(1))
```


# Explore the emperical data
```{r}
# Mean of correct by participant
df %>% 
  group_by(subject) %>% 
  summarize(mean(correct, na.rm = TRUE))

# Overall mean of correct
df %>% 
  ungroup() %>% 
  summarize(mean(correct, na.rm = TRUE))

# Performance over time 

## Add performance to the df
df <- df %>%
  group_by(subject) %>% 
  mutate(performance = cumsum(correct) / seq_along(correct))

## Per participant
ggplot(df, aes(x = trial, y = performance, color = as.factor(subject))) +
  geom_line() +                
  geom_point() +               
  labs(title = "Performance per subject",
       x = "Trial",
       y = "Performance",
       color = "Subject") +    
  theme_minimal()  
```

# Reduce the dimensions in the real data (since we are not using "nutritious")

Responses are 1-4, where 3 and 4 indicate danger, 1 and 2 not danger; 2 and 4 indicate nutritious, 1 and 3 not nutritious

Category: the category to which the alien belongs. 1: peaceful/non-nutritious, 2: peaceful/nutritious, 3: dangerous/non-nutritious, 4: dangerous/nutritious.

```{r}
# Reduce from 2 dimension (dangerous/nutritious) to 1 dimension (dangerous)
df <- df %>% 
  dplyr::select(-nutricious, -category) %>% 
  mutate(response_new = case_when(
    response == 1 ~ 0,
    response == 2 ~ 0,
    response == 3 ~ 1,
    response == 4 ~ 1)) %>% 
  mutate(correct_new = ifelse(response_new == dangerous, 1, 0)) # if response_new and dangerous is the same put a 1 in the correct_new, otherwise a 0

# Move the columns around
df <- df %>% 
  relocate(response_new, .after = response) %>% 
  relocate(correct_new, .after = correct)
```


# Explore the emperical data again after the dimension reduction
```{r}
# Mean of correct by participant
df %>% 
  group_by(subject) %>% 
  summarize(mean(correct_new, na.rm = TRUE))

# Overall mean of correct
df %>% 
  summarize(mean(correct_new, na.rm = TRUE))

# Performance over time 

## Add performance to the df
df <- df %>%
  group_by(subject) %>% 
  mutate(performance_new = cumsum(correct_new) / seq_along(correct_new))

## Per participant
ggplot(df, aes(x = trial, y = performance_new, color = as.factor(subject))) +
  geom_line() +                
  geom_point() +               
  labs(title = "Performance per subject",
       x = "Trial",
       y = "Performance",
       color = "Subject") +    
  theme_minimal()  
```

# Limit to 5 participants 
```{r}
df_5 <- df %>% 
  filter(subject %in% c(1, 2, 3,4,5))

```


# Loop over 5 participants and apply STAN code
```{r}
# Specify path to stan model and load 
file <- file.path("gcm_stan.stan")

model <- cmdstan_model(file, 
                     cpp_options = list(stan_threads = TRUE),
                     stanc_options = list("O1"))

# Get the unique subject IDs 
subject_ids <- unique(df_5$subject)

# Store model fits 
model_fits_emp <- list()

for (subj in subject_ids) {
  cat("Fitting model for subject:", subj, "\n")
  
  subj_data <- df_5 %>% filter(subject == subj)
  
  data <- list(
    ntrials = nrow(subj_data),
    nfeatures = 5,
    cat_one = subj_data$dangerous,
    y = subj_data$response_new,
    obs = as.matrix(subj_data[, c("F1_eyes", "F2_legs", "F3_body", "F4_arms", "F5_color")]),
    b = 0.5,
    w_prior_values = rep(1, 5),
    c_prior_values = c(0, 1)
  )
  
  fit <- model$sample(
    data = data,
    seed = 123,
    chains = 2, 
    parallel_chains = 2, 
    threads_per_chain = 2, 
    iter_warmup = 1000,
    iter_sampling = 1000,
    refresh = 0,
    max_treedepth = 20,
    adapt_delta = 0.99
  )
  
  model_fits_emp[[as.character(subj)]] <- fit
}


# Print diagnostics 
for (subj in names(model_fits_emp)) {
   cat("\n--- Diagnostics for Subject", subj, "---\n")
   print(model_fits[[subj]]$cmdstan_diagnose())
}

# Print and save diagnostics per subject (to be used later)
for (subj in names(model_fits_emp)) {

  #Extract summary and convert to tibble
  diagnostics <- model_fits_emp[[subj]]$summary() %>% as_tibble()
  
  # Assign subject name 
  assign(paste0("subject_", subj, "_diagnostics_empirical"), diagnostics)
  
  # Save the tibble as an RDS file
  saveRDS(diagnostics, paste0("subject_", subj, "_diagnostics.rds"))
  
  # Print the summary
  print(diagnostics)
}

# Plot traceplots
for (subj in names(model_fits_emp)) {
  
  # Extract posterior draws
  draws <- as_draws_df(model_fits_emp[[subj]]$draws())
  
  # Prior-posterior update checks for w's
  w_prior_pos <- lapply(1:5, function(i) {
    ggplot(draws)+
      geom_density(aes(x = .data[[paste0("w[", i, "]")]]), alpha = 3, fill = "lightblue") +
      geom_density(aes(x = .data[[paste0("w_prior[", i, "]")]]), alpha = 3, fill = "pink") +
      geom_vline(xintercept = 0.2, size = 1) + 
      theme_classic() + 
      ggtitle(paste("Prior-Posterior Update for w[", i, "]", sep = "")) +
      xlab(paste0("w[", i, "]")) +
      ylab("Posterior Density")
  })
  
  # Prior-posterior update checks for c
  c_prior_pos <- ggplot(draws) +
    geom_density(aes(c), alpha = 3, fill = "lightblue") +
    geom_density(aes(c_prior), alpha = 3, fill = "pink") +
    geom_vline(xintercept = 1, size = 1) + 
    theme_classic() + 
    ggtitle("Prior-Posterior Update for c") + 
    xlab("c") +
    ylab("Posterior Density")
  
  # Combine the prior-posterior update checks
  grid_plots <- c(w_prior_pos, list(c_prior_pos))
  grid.arrange(grobs = grid_plots, 
               ncol = 2, 
               top = paste("Prior-posterior update checks - Subject", subj),
               bottom = "Prior = pink, Posterior = blue")
  
  # Traceplots for w's
  w_plots <- lapply(1:5, function(i) {
    ggplot(draws, aes(x = .iteration, y = .data[[paste0("w[", i, "]")]], 
                      group = .chain, color = .chain)) +
      geom_line(alpha = 0.5) +
      theme_classic() +
      ggtitle(paste("w[", i, "]", sep = ""))
  })
  
  # Traceplot for c 
  c_plot <- ggplot(draws, aes(x = .iteration, y = c, group = .chain, color = .chain)) +
    geom_line(alpha = 0.5) +
    theme_classic() +
    ggtitle("c")
  
  # Combine 
  grid_plots <- c(w_plots, list(c_plot))
  grid.arrange(grobs = grid_plots, ncol = 2, top = paste("Traceplots - Subject", subj))
}

```



# Combine the diagnostic outputs
```{r}
# Get the data frames into the right shape
subj_1_empirical <- subject_1_diagnostics_empirical %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)

subj_2_empirical <- subject_2_diagnostics_empirical %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)

subj_3_empirical <- subject_3_diagnostics_empirical %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)

subj_4_empirical <- subject_4_diagnostics_empirical %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)

subj_5_empirical <- subject_5_diagnostics_empirical %>% 
  filter(variable == "lp__"|variable == "w[1]"|variable == "w[2]"|variable == "w[3]"|variable == "w[4]"|variable == "w[5]"|variable == "c") %>% 
  dplyr::select(variable, mean, median, sd, mad)


# Combine the data frames
combined_data_empirical <- bind_rows(subj_1_empirical, subj_2_empirical, subj_3_empirical, subj_4_empirical, subj_5_empirical)


# Calculate the mean for each variable across participants
mean_data_empirical <- combined_data_empirical %>%
  group_by(variable) %>%
  summarise(mean_value = mean(mean),
            median_value = median(mean),
            sd_value = sd(mean),
            mad_value = mad(mean))

mean_data_empirical
```

